Ignition of Fuel Beds by Firebrands

### 2022-05-30 conclusions.tex
Updates to ignition model

\section{Sensitivities of Porous Beds and Plates to Ignition by Firebrands}
        \begin{enumerate}
            \item 
                Smaller particles ignite more readily in porous beds than larger particles when heat transfer from the heater is primarily through conduction. This was evident by higher ignition probabilities, in general, of the smaller particles for a fixed heater temperature. As particle sizes increase radiant heat transfer becomes more important and fuel beds with larger particles were more likely than smaller particles to ignite at extended times (\textgreater 100\si{\second}) due to the increased importance of radiant ignition. 
            \item
                Douglas-fir plates ignite at times where conduction is the dominant mode of heat transfer (\textless 10\si{\second}) due to the higher thermal conductivity of the solid plates. The ignition probability of plates was the most similar to the larger particle, in particular at lower heater temperatures, due to dispersed heating of the porous fuel bed through radiation and the increased thermal conductivity of the plates creating similar temperature profiles. The rise in ignition probability  over a smaller heater temperature range time with temperature results from more consistent contact between the heater and plate surface.
            \item 
                Heat flux delivered to the fuel bed, when compared to heater temperature, is more indicative of ignition likelihood and ignition time for porous fuel beds. Heat flux is a more significant predictor of ignition because it captures differences in heat transfer modes and particle contact that heater temperature values do not. While this finding is not new, what is novel is that the mixed mode of heating (conduction and radiation) has a significant impact on the flaming ignition of fuel beds.
            \item 
                Consideration of the transport characteristics of pyrolyzate gases near the high temperature source can be important for more fully predicting ignition propensity. A \textit{Da} of ignition, in relation to the measured heat flux and thermal diffusivity of the fuel beds, is a promising relationship for predicting ignition for the porous fuel beds.  
        \end{enumerate}

@@

\section{Effect of Fuel Bed Composition on Flaming Ignition Probability}
        \begin{enumerate}
            \item Of the six materials tested, only three were able to be ignited within below the 750\si{\celsius} maximum temperature of the firebrand surrogate. Two of the three materials for which ignition was not observed were bark. The higher ignition threshold of the bark materials is attributed to higher lignin concentrations. The resistance to ignition present in these materials suggest that using materials high in lignin in locations at risk for firebrand attack in wildfires may reduce ignitions and subsequent losses of structures.

@@

\section{Influence of Multiple Firebrands on the Ignition of Fuel Beds}
        \begin{enumerate}
            \item An increase of wind speed from 0.5\si{\meter\per\second} to 5.8\si{\meter\per\second} reduced the temperature required for ignition for all heater configurations. The decrease in ignition temperature required ranged from 20\% to 60\% depending on the heater configuration.

            \item At wind speeds of 5.8\si{\meter\per\second} the ignition threshold is largely independent of the heater configuration. This lack of sensitivity is attributed to the ignition being controlled by the fluid dynamics around the firebrand surrogates. Under wind, ignition is largely controlled by the propensity of pyrolysis products to accumulate in recirculation zones and independent of thermal interactions with nearby objects, whether they are energy sources or sinks. 
            \item At wind speeds of 0.5\si{\meter\per\second} the ignition threshold is dependent on the firebrand surrogate configuration. For example, when the firebrands were spaced one diameter apart, the difference between ignition thresholds when the upstream firebrand is inert or an energy source is 34\%. When the firebrand surrogates were spaced such that they did not thermally interact (five diameters apart), the difference in ignition threshold was 4\%. Thus, at low wind speeds or in quiescent conditions, ignition is sensitive to thermal interactions with nearby objects. Ignition is promoted if nearby objects supply energy and inhibited if energy sinks are nearby.

            \item In configurations where the firebrand surrogate is downstream of an inert flow obstacle, ignition may occur as a smoldering to flaming transition. The smoldering to flaming transition is facilitated by either accumulation of pyrolysis gases in the downstream edge of the flow obstruction or by forming an overhang as the fuel bed recedes underneath the obstruction. This phenomenon produces ignition at similar temperatures to other configurations and may not be important for ignition predictions but may be of interest to smoldering research. 
        \end{enumerate}
    The conclusions of this work show that ignition of a fuel bed is more likely when a firebrand or firebrand lands on a fuel bed under conditions that promote recirculation zones near locations where pyrolysis gases are released from the fuel bed. It was observed that under windy conditions ignition is less sensitive to nearby embers or inert objects than at low wind speeds. These observed sensitivities highlight the increased risk of fuel bed ignition in windy conditions and in cases where firebrands may accumulate in all wind conditions.

@@

    The aggregated data was then used to create a Random Forest Classifier model using the scikit-learn package in python~\cite{scikit-learn} to estimate ignition outcomes for each test. Using a 0.25/0.75 test train spit on the data, the R\textsuperscript{2} validation score for the model with an out-of-bag score of 0.83, which means that the model predicts ignition correctly 83\% of the time. While the accuracy of the model is likely acceptable if such a model was to be used as an indicator or predictor of ignition during a fire or to inform preventative measures, it would be beneficial to reduce the number of inputs required. An examination of the most important input parameters and refinement of the model found that accuracy was not degraded if the number of parameters was reduced to include only the following variables:

@@

    Reducing the model to include only the five characteristics above maintained the 83\% prediction accuracy. While it is beneficial to have an accurate predictive model for a single set of experiments in a similar apparatus, a model capable of predictions across different is far more valuable for preventing home loss. The model was applied to select studies from the literature to evaluate the accuracy across studies. The authors and relevant parameters from each study are shown in Table~\ref{tab:otherStudies}. All of the studies presented in Table~\ref{tab:otherStudies} used hot metal particles dropped onto fine fuel beds of processed material. The studies by Hadden et al. and Urban et al. used pure cellulose as the fuel bed, and the study by Zak et al. used a blend of grasses processed into a powder. A wind direction of 0\si{\degree} was used for all studies since the geometry of the metal particles is anticipated to produce recirculation zones similar to that of a cylindrical heater parallel to the flow. The average temperature was defined as the average of the initial temperature and ambient temperature. More sophisticated estimates of the average temperature based on the average time to ignition and estimated heat losses from the ember were conducted, but the model accuracy was not improved. Thus, a simpler metric was chosen.  

@@

                 Zak et al.    & 310 - 560 & 338 & 0.5 & 0.04 - 0.15 & 0 & \cite{Zak2014}

@@

    The first approach used was to apply a reduced model created from only the data generated from studies in this dissertation and predict ignition for each of the studies in the literature. The accuracy of the model at predicting ignition for other studies was 50\%, which is entirely unusable for any type of predictive modeling. However, when the model was retrained using the same five parameters but including the Zak et al. data in conjunction with the results of this work, the predictive accuracy increased to 96\% for the Hadden et al. study and 81\% for the Urban et al. study. The increase in accuracy suggests that predictions can be made across differences in experimental conditions using as few as five parameters. Predictions across experimental conditions have been one of the biggest hurdles to creating a general predictive model. Identifying the primary controlling parameters of ignition in this work has created a framework for bridging the predictive gap across various configurations.

>>>>

\section{Sensitivities of Porous Beds and Plates to Ignition by Firebrands}
        \begin{enumerate}
            \item 
                Smaller particles ignite more readily in porous beds than larger particles when heat transfer from the heater is primarily through conduction. This was evident by higher ignition probabilities, in general, of the smaller particles for a fixed heater temperature. As particle sizes increase radiant heat transfer becomes more important and fuel beds with larger particles were more likely than smaller particles to ignite at extended times (\textgreater 100\si{\second}) due to the increased importance of radiant ignition. 
            \item
                Douglas-fir plates ignite at times where conduction is the dominant mode of heat transfer (\textless 10\si{\second}) due to the higher thermal conductivity of the solid plates. The ignition probability of plates was the most similar to the larger particle, in particular at lower heater temperatures, due to dispersed heating of the porous fuel bed through radiation and the increased thermal conductivity of the plates creating similar temperature profiles. The rise in ignition probability  over a smaller heater temperature range time with temperature results from more consistent contact between the heater and plate surface.
            \item 
                Heat flux delivered to the fuel bed, when compared to heater temperature, is more indicative of ignition likelihood and ignition time for porous fuel beds. Heat flux is a more significant predictor of ignition because it captures differences in heat transfer modes and particle contact that heater temperature values do not. While this finding is not new, what is novel is that the mixed mode of heating (conduction and radiation) has a significant impact on the flaming ignition of fuel beds.
            \item 
                Consideration of the transport characteristics of pyrolyzate gases near the high temperature source can be important for more fully predicting ignition propensity. A \textit{Da} of ignition, in relation to the measured heat flux and thermal diffusivity of the fuel beds, is a promising relationship for predicting ignition for the porous fuel beds.  
        \end{enumerate}
    Further work is needed to verify that the \textit{Da} may be used to predict ignition for solid surfaces and for porous fuel beds with varying chemical compositions.  If proven valid, the (\textit{Da}), measured/predicted heat fluxes, and fuel bed properties may be used to help predict ignition of fuel beds both in and out of the WUI, ultimately helping to increase the effectiveness of fire prevention and suppression efforts.

@@

\section{Effect of Fuel Bed Composition on Flaming Ignition Probability}
        \begin{enumerate}
            \item Of the six materials tested, only three were able to be ignited within below the 750\si{\celsius} maximum temperature of the firebrand surrogate. Two of the three materials for which ignition was not observed were bark. The higher ignition threshold of the bark materials is attributed to higher lignin concentrations. The resistance to ignition present in these materials suggest that using materials high in lignin in locations at risk for firebrand attack in wildfires may reduce ignitions and subsequent losses of structures. It is acknowledged that these results are for flaming ignition only, and high lignin materials may pose a significant risk if smoldering or the smoldering to flaming transition occurs.

@@

\section{Influence of Multiple Firebrands on the Ignition of Fuel Beds}
        \begin{enumerate}
            \item Increasing the wind speed lowers the threshold for ignition. However, the magnitude of the change in ignition threshold is sensitive to the spacing and temperatures of cartridge heaters used. The ignition threshold may reduce between 20\% and 60\% depending on the heater configurations.

            \item At wind speeds of 5.8\si{\meter\per\second} the ignition threshold is largely independent of thermal interactions between firebrand surrogates. The lack of sensitivity to inter-firebrand effects is attributed to the ignition being largely controlled by the fluid dynamics around the firebrand surrogates. Ignition is largely controlled by the propensity of pyrolysis products to accumulate and is independent of thermal interactions with nearby objects, whether they are energy sources or sinks.

            \item In low wind conditions (e.g., 0.5\si{\meter\per\second}), the ignition threshold is more sensitive to thermal interactions between the firebrand surrogates than in higher wind conditions. For example, when the firebrand was spaced one diameter apart, the difference between ignition thresholds when the upstream firebrand is inert or an energy source is 34\%. When the firebrand surrogates were spaced such that they did not thermally interact (five diameters apart), the difference in ignition threshold was 4\%. Thus, at low wind speeds or in quiescent conditions, ignition is sensitive to thermal interactions with nearby objects. Ignition is promoted if nearby objects supply energy and inhibited if thermal sinks are nearby.

            \item In configurations where the firebrand is downstream of a flow obstacle, ignition may occur via the smoldering to flaming transition. The smoldering to flaming transition appears to be facilitated by either accumulation of pyrolysis gases in the downstream edge of the flow obstruction or by the formation of an overhang as the fuel bed recedes underneath the obstruction. This phenomenon produces ignition at similar temperatures to other configurations and may not be important for ignition predictions but may be of interest to smoldering research. 
        \end{enumerate}
    The conclusions of this work show that ignition of a fuel bed is more likely when a firebrand or firebrand lands on a fuel bed under conditions that promote recirculation zones near locations where pyrolysis gases are released from the fuel bed. It was observed that under windy conditions ignition is less sensitive to nearby embers or inert objects than at low wind speeds. These observed sensitivities highlight the increased risk of fuel bed ignition in windy conditions and in cases where firebrands may accumulate in all wind conditions.

@@

    The aggregated data was then used to create a Random Forest Classifier model using the scikit-learn package in python~\cite{scikit-learn} to estimate ignition outcomes for each test. A 0.75/0.25 test-train split was used for training the models, meaning that 25\% of the data was used to train the model and the remaining 75\% was used to test the model accuracy. The R\textsuperscript{2} validation score for the model with an out-of-bag score of 0.83, which means that the model predicts ignition correctly 83\% of the time. While the accuracy of the model is likely acceptable if such a model was to be used as an indicator or predictor of ignition during a fire or to inform preventative measures, it would be beneficial to reduce the number of inputs required. An examination of the most important input parameters and refinement of the model found that accuracy was not degraded if the number of parameters was reduced to include only the following variables:

@@

    Reducing the model to include only the five characteristics above maintained the 83\% prediction accuracy. While it is beneficial to have an accurate predictive model for a single set of experiments in a similar apparatus, a model capable of predictions across different is far more valuable for preventing home loss. The model was applied to select studies from the literature to evaluate the accuracy across studies. The authors and relevant parameters from each study are shown in Table~\ref{tab:otherStudies}. Three of the studies presented in Table~\ref{tab:otherStudies} used hot metal particles dropped onto fine fuel beds of processed material. The studies by Hadden et al. and Urban et al. used pure cellulose as the fuel bed, and the study by Zak et al. used a blend of grasses processed into a powder. The fourth study, conducted by Filkov et al., used multiple bark embers of various sizes dropped onto dried, but otherwise unprocessed, pine needle duff. A wind direction of 0\si{\degree} was used for for the not metal particle studies since the geometry of the metal particles is anticipated to produce recirculation zones similar to that of a cylindrical heater parallel to the flow. A wind direction if 45\si{\degree} was used for the pine bark embers since the embers provided a more significant flow disruption than the metal particles but were dropped randomly and therefore not considered to be normal to the flow in all cases. The average temperature was defined as the average of the initial temperature and ambient temperature for the hot metal particles. More sophisticated estimates of the average temperature based on the average time to ignition and estimated heat losses from the ember were conducted, but the model accuracy was not improved. Thus, a simpler metric was chosen. The average temperature of the bark embers was estimated from measurements of similar embers in the literature~\cite{Fateev2017a, Matvienko2018}.

@@

                 Zak et al.    & 310 - 560 & 338 & 0.5       & 0.04 - 0.15 & 0  & \cite{Zak2014} \\
                 Filkov et al. & 407       & 105 & 1, 1.5, 2 & 0.17 - 2.60 & 45 & \cite{Filkov2016}

@@

    The first approach used was to apply a reduced model created from only the data generated from studies in this dissertation and predict ignition for each of the studies in the literature. The accuracy of the model at predicting ignition for other studies was 50\%, which is entirely unusable for any type of predictive modeling. However, when the model was retrained using the same five parameters but including 25\% subsets of the Zak et al. and the Filkov et al. data, in conjunction with the results of this work, the predictive accuracy increased to 98\% for the Zak et al. study, 94\% for the Hadden et al. study, 81\% for the Urban et al. study and 74\% for the Filkov et al. study. The increase in accuracy suggests that predictions can be made across differences in experimental conditions, ember types, and fuel bed types using as few as five parameters. Predictions across experimental conditions have been one of the biggest hurdles to creating a general predictive model. Identifying the primary controlling parameters of ignition in this work has created a framework for bridging the predictive gap across various configurations.

### 2022-05-30 futureWork.tex
Updates to ignition model

  The results in this work have identified controlling parameters for ignition and provided a framework for creating models with capabilities to predict various ignition scenarios. Significant additional work is needed to refine the proposed framework and explore different ignition parameters. 
        \begin{itemize}
            \item The moisture content of the fuel bed was not considered in this work but is a well-studied parameter that significantly affects ignition. Extension of the experiments is necessary to determine how moisture content fits into the proposed framework. The change in fuel bed bulk density due to moisture may sufficiently characterize the difference, but that remains to be seen.

            \item Additional tests with well-characterized smoldering or flaming embers would significantly enhance the knowledge gained in this work. Typical temperatures of physically combusting firebrands are higher than those capable by the apparatus used in this work~\cite{Fateev2017a}. The higher temperatures may influence the ignition propensity. However, combusting embers are susceptible to heat losses and environmental factors where the firebrand surrogates used in this work are mainly independent of these factors. It is unclear how or if the addition of energy sources that react with the environment will shift the parameters identified to be most influential to ignition. A series of highly instrumented and controlled experiments focused on the coupled interactions between a reacting ember, a reacting fuel bed, and changing environmental conditions. I applaud the patience and perseverance of those who may undertake this endeavor in the future.

@@

            \item In addition to the limited data available for thermal properties of fuel bed materials, characterization of the chemical composition of materials and the influence of chemical changes on the pyrolysis and subsequent ignition would provide valuable insight into the differences in ignition between materials and environmental conditions. 
        \end{itemize}

>>>>

  The results in this work have identified controlling parameters for ignition and provided a framework for creating models with capabilities to predict various ignition scenarios. Significant additional work is needed to refine the proposed framework and explore different ignition parameters. 
        \begin{itemize}
            \item The moisture content of the fuel bed was not considered in this work but is a well-studied parameter that significantly affects ignition. Extension of the experiments, or a meta-analysis of literature is necessary to determine how moisture content fits into the proposed framework. The change in fuel bed bulk density due to moisture may sufficiently characterize the difference, but that remains to be seen.

            \item Additional tests with well-characterized smoldering or flaming embers would significantly enhance the knowledge gained in this work. Typical temperatures of physically combusting firebrands are higher than those capable by the apparatus used in this work~\cite{Fateev2017a}. The higher temperatures may influence the ignition propensity. However, combusting embers are susceptible to heat losses and environmental factors where the firebrand surrogates used in this work are mainly independent of these factors. It is unclear how or if the addition of energy sources that react with the environment will shift the parameters identified to be most influential to ignition. A series of highly instrumented and controlled experiments focused on the coupled interactions between a reacting ember, a reacting fuel bed, and changing environmental conditions would provide novel and valuable insight into ignition processes. I applaud the patience and perseverance of those who may undertake this endeavor in the future.

@@

            \item In addition to the limited data available for thermal properties of fuel bed materials, the chemical composition of materials and the influence of chemical changes on the pyrolysis and subsequent ignition would provide valuable insight into the differences in ignition between materials and environmental conditions.

        \end{itemize}

### 2022-05-30 manuscript3.tex
Updates to ignition model

\section{Conclusions}
\label{sec:conclusions3}    
    Flaming ignition tests were conducted for fuel beds of six different materials at two different wind speeds. Ignition of fuel beds was induced by a cartridge heater that served as a firebrand surrogate. The temperature of the heater was maintained at a fixed temperature throughout each test allowing sensitivities of ignition to fuel bed composition to be evaluated. The ignition probability and chemical composition were used to evaluate the influence of chemical composition of the fuel bed and observe differences in ignition probabilities in the presence of wind and in quiescent conditions. The specific conclusions of this work are as follows:
        \begin{enumerate}
            \item Of the six materials tested, only three were able to be ignited within below the 750\si{\celsius} maximum temperature of the firebrand surrogate. Two of the three materials for which ignition was not observed were bark. The higher ignition threshold of the bark materials is attributed to higher lignin concentrations. The resistance to ignition present in these materials suggest that using materials high in lignin in locations at risk for firebrand attack in wildfires may reduce ignitions and subsequent losses of structures. It is acknowledged that these results are for flaming ignition only and high lignin materials may pose significant risk if smoldering or the smoldering to flaming transition occurs.

>>>>

\section{Conclusions}
\label{sec:conclusions3}    
    Flaming ignition tests were conducted for fuel beds of six different materials at two different wind speeds. Ignition of fuel beds was induced by a cartridge heater that served as a firebrand surrogate. The temperature of the heater was maintained at a fixed temperature throughout each test, allowing sensitivities of ignition to fuel bed composition to be evaluated. The ignition probability and chemical composition were used to evaluate the influence of the chemical composition of the fuel bed and observe differences in ignition probabilities in the presence of wind and in quiescent conditions. The specific conclusions of this work are as follows:
        \begin{enumerate}
            \item Of the six materials tested, only three were able to be ignited within below the 750\si{\celsius} maximum temperature of the firebrand surrogate. Two of the three materials for which ignition was not observed were bark. The higher ignition threshold of the bark materials is attributed to higher lignin concentrations. The resistance to ignition present in these materials suggest that using materials high in lignin in locations at risk for firebrand attack in wildfires may reduce ignitions and subsequent losses of structures. It is acknowledged that these results are for flaming ignition only, and high lignin materials may pose a significant risk if smoldering or the smoldering to flaming transition occurs.
